---
title: "assignment_5_blaufuss"
author: "Dennis Blaufuss"
date: "12/10/2021"
output:
  html_document:
    df_print: paged
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(rethinking)
library(dagitty)

# Note: I knit to html and then save the html file to pdf to receive the submitted result. Thus this Markdown isn't optimized for knitting to pdf. I choose this workaround since I encountered a lot of weird issues with knitting the pdf in the first place (I guess M1 related as usual :D)
```

##### For this assignment, you will work with the Trolley dataset:

```{r}
data(Trolley)
d <- Trolley
precis(d)
```

##### The basis of the assignment are the models developed in sections 12.3-12.4 of the textbook. So, you should first implement that code before starting this problem set.

```{r}

```

***

##### 1. We see that education, modeled as an ordered category, is associated with moral judgments. Is this association causal? One possible confound is that education is also associated with age through a causal process: namely, people are older when they finish a level of education than when they begin it.
##### Reconsider the Trolley data in light of this issue. Specifically,
##### (a) Draw a DAG that represents hypothetical causal relationships among response, education, and age.
##### (b) Identify which statistical model or models are required to evaluate the causal influence of education on responses. Hint: you should use the code in the book to reorder the education level labels:
```{r}
# R code 12.31
edu_levels <- c( 6 , 1 , 8 , 4 , 7 , 2 , 5 , 3 )
d$edu_new <- edu_levels[ d$edu ]
```
##### (c) What do you conclude about the causal relationships among these three variables?

###### Answer:

```{r}

```

***

##### 2. Now consider one more variable in the Trolley data: Gender. Suppose that gender might influence education as well as response directly. Draw the DAG now that in- cludes response, education, age, and gender. Is it possible that any of the inferences from Question 1 are confounded by gender? If so, define any additional models you need to infer the causal influence of education on response. What do you conclude?

###### Answer:

```{r}

```

Text

***

##### 3. Rewrite the following model as a multilevel model.
yi ∼ Binomial(1, pi) 
logit(pi) = αgroup[i] +βxi
αgroup ∼ Normal(0, 1.5) 
β ∼ Normal(0, 0.5)

###### Answer:
